```{r echo=FALSE, results='hide', message=FALSE}
library(ggplot2)
```
---
title: "Statistical Inference Course Project"
output: word_document
---

## Assignment

In this project you will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. You will investigate the distribution of averages of 40 exponentials. Note that you will need to do a thousand simulations.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials.  You should
1. Show the sample mean and compare it to the theoretical mean of the distribution.
2. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
3. Show that the distribution is approximately normal.

In point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials. 

## Overview
This task demonstrates the central limit theorem, which states
the distribution of the sample means follows the 
central limit theorem, with states:
* mean of sample means = xbar = population mean
* sd of sample means = s = population mean/sqrt(n)
  * where n is the sample size
  * and sample means are normally (gaussian) distributed
regardless of the population distribution, exponential in this case

We do this by obtaining 1,000 samples of sample size 40 from a population
exponentially distributed with lambda (rate) of 5. According to
[Wolfram MathWorld](http://mathworld.wolfram.com/ExponentialDistribution.html),
given a Poisson distribution with rate of change lambda, the distribution of waiting times between successive changes (with k=0) is exponentially
distributed with mean 1/lambda and standard deviation of 1/lambda.

Therefore, according to the central limit theorem, the distribution of 
the means of the 1,000 samples should be normally distributed with mean
1/lambda and standard deviation (1/lambda)/sqrt(40).

With a lambda of 0.2, this translates to 
mean = 5 and 
sd = 5/6.25 = 0.8.

## Simulations
: Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.

The following code generates 1,000 samples of n=40 from an exponential 
distribution with lambda = 0.2
``` {r}
lambda <- rep(0.2, 1000)
sample_size <- 40
sim <- lapply(lambda, function(x) rexp(sample_size, x))
```
The following code calculates the means and standard deviations of the sample means.
```{r} 
means <- sapply(sim, mean) # vector of length 1,000 (length(means))
mean(means) # approx 1/lambda = 1/0.8 = 5
sd(means) # approx 5/sqrt(40) = 5/6.25 = 0.8

sds <- sapply(sim, sd)
vars <- sapply(sim, var)
mean(sds)
sd(sds)
```

## Sample Mean versus Theoretical Mean
: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.
```{r echo=FALSE, results='hide', message=FALSE}
# Function factory. DRY.
plotFunction <- function(fun, pl=seq(2,8)) {
        
        function(histx, argslist, plotlimits=pl) {
                df <- data.frame(vals=histx)
                print(args)
                ggplot() + 
                        geom_histogram(data = df, 
                                       mapping = aes(x=vals, y=..density..)) + 
                        stat_function(data = data.frame(x=plotlimits), 
                                      mapping = aes(x), 
                                      fun = fun, # defines the generated function
                                      args = argslist, # must accommodate different dist'ns
                                      color = "red")
        }
}
```


NEED TO SHOW THE DISTRIBUTION OF THE SAMPLE MEAN AND THE 95%
CONFIDENCE INTERVAL. SHOW THE POPULATION MEAN (KNOWN) FALLS WITHIN
THE 95% CONFIDENCE INTERVAL. 
Maybe even show the forumula in mathematical notation.

```{r}
# normPlot <- function(histx, mu, sigma, plotlimits=seq(2,8)) {
#         df <- data.frame(vals=histx)
#         ggplot() + 
#                 geom_histogram(data = df, 
#                                mapping = aes(x=vals, y=..density..)) + 
#                 stat_function(data = data.frame(x=plotlimits), 
#                               mapping = aes(x), 
#                               fun = dnorm, 
#                               args = list(mean=mu, sd=sigma), 
#                               color = "red")
# }
theoretical_mean_of_means = 1/0.2
theoretical_sd_of_means = (1/0.2)/sqrt(sample_size)

normPlot <- plotFunction(dnorm)
normPlot(means, list(mean=theoretical_mean_of_means, sd=theoretical_sd_of_means))
```

## Sample Variance versus Theoretical Variance
: Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.

NEED TO SHOW THE DISTRIBUTION OF THE SAMPLE VARIANCE AND THE 95%
CONFIDENCE INTERVAL. SHOW THE POPULATION VARIANCE (KNOWN) FALLS WITHIN
THE 95% CONFIDENCE INTERVAL.
Maybe even show the forumula in mathematical notation.

```{r}
# expPlot <- function(histx, rate, plotlimits=seq(2,8)) {
#         dataframe <- data.frame(vals=histx)
#         ggplot() + 
#                 geom_histogram(data = dataframe, 
#                                mapping = aes(x=vals, y=..density..)) + 
#                 stat_function(data = data.frame(x=plotlimits), 
#                               mapping = aes(x), 
#                               fun = dexp, 
#                               args = list(rate=rate), 
#                               color = "red")
# }

theoretical_mean_of_sds = 1/0.2
theoretical_sd_of_sds = 1 # (1/0.2)/sqrt(sample_size)

expPlot <- plotFunction(dexp)
expPlot(sds[[10]], list(rate=0.2)) # underlying dist'n is exp

# chisqPlot <- function(histx, df, plotlimits=seq(2,8)) {
#         dataframe <- data.frame(vals=histx)
#         ggplot() + 
#                 geom_histogram(data = dataframe, 
#                                mapping = aes(x=vals, y=..density..)) + 
#                 stat_function(data = data.frame(x=plotlimits), 
#                               mapping = aes(x), 
#                               fun = dchisq, 
#                               args = list(df=df), 
#                               color = "red")
# }

chisqPlot <- plotFunction(dschisq)
chisqPlot(x, list(df=39)) # this is nothing yet
```

## Distribution
: Via figures and text, explain how one can tell the distribution is approximately normal. 


